---
title: "hw_1"
author: "Danielle Sclafani"
date: "4/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(chron)
library(here)
library(janitor)
```

#### Reading in the data
```{r}
climate_data <- read_csv(here("data", "bishop_airport.csv")) %>% 
  clean_names()
```

#### visualizing the data
```{r}

ggplot(climate_data, aes(date, climate_data$tmax))+geom_line()+ labs(y="Daily Maximum Temperature (degrees F)", x="Date")

ggplot(climate_data, aes(date, climate_data$tmin))+geom_line()+ labs(y="Daily Minimum Temperature (degrees F)", x="Date")

ggplot(climate_data, aes(date, climate_data$prcp))+geom_line()+ labs(y="Daily Rainfall (in)", x="Date")

```

```{r}
#missing a big chunk of data between 1898, going to choose data between 1950-2021, because missing large chunk before 1950

clim_1950 <- climate_data %>% 
  mutate(year = year(date)) %>% 
  filter(year >= 1950,
         year != 2021) #removed 2021 because do not have a full year of data, was skewing data

# need to fill the rows that are N/As
fillrow = which(is.na(clim_1950$prcp))
fillrow
clim_1950$prcp[fillrow]=0

# find rows with missing data
# temperature
fillrow = which(is.na(clim_1950$tmax) | clim_1950$tmax < 40)
fillrow = fillrow[2:length(fillrow)]
clim_1950$tmax[fillrow]=(clim_1950$tmax[fillrow+1]+clim_1950$tmax[fillrow-1])/2

fillrow = which(is.na(clim_1950$tmin) | clim_1950$tmin < 40)
fillrow = fillrow[2:length(fillrow)]
clim_1950$tmin[fillrow]=(clim_1950$tmin[fillrow+1]+clim_1950$tmin[fillrow-1])/2
```

```{r}
#replotting ot make sure this works
ggplot(clim_1950, aes(date, tmax))+geom_line()+ labs(y="Daily Maximum Temperature (degrees F)", x="Date")

ggplot(clim_1950, aes(date, tmin))+geom_line()+ labs(y="Daily Minimum Temperature (degrees F)", x="Date")

ggplot(clim_1950, aes(date, prcp))+geom_line()+ labs(y="Daily Rainfall (in)", x="Date")

```


The trend has a lot of seasonality and is very noisy so going to try annual averages - adding a trend line to the tmin and tmax
```{r, fig.cap="Figure 1.0 Initial Data Visualization of Annual Average Daily Minimum and Maximum temperatures from 1950-2020"}
clim_year_avg <-
  clim_1950 %>% 
  group_by(year) %>% 
  summarize(tmax_mean=mean(tmax), tmin_mean = mean(tmin), precip_mean=mean(prcp))


a <- ggplot(data = clim_year_avg, aes(x = year, y = tmax_mean)) + geom_point(color = "skyblue") +
  labs(x = "Year", y = "Average Temperature (degrees F)", title = "Average Annual Minimum and Maximum Temperatures (degrees F)") +
  scale_y_continuous(limits=c(min(clim_year_avg$tmin_mean), max(clim_year_avg$tmax_mean)))+
geom_point(data=clim_year_avg, aes(x=year, y = tmin_mean), col="cyan4")

#adding a trendline
a = a+stat_smooth(method = "lm", col = "skyblue")
a
a+stat_smooth(data = clim_year_avg, aes(x = year, y = tmin_mean), col = "cyan4", method = lm)+theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5), text = element_text(family = "Times New Roman"))

ggsave(here("figs", "yearavg_min_max_temps.png"))
```


## Finding the slope of the trendlines
```{r}

# regression line for min temp, year average

res = lm(tmin_mean~year, data=clim_year_avg)
summary(res)
confint(res, "year", level = 0.95)
ggplot(clim_year_avg, aes(x = year, y = tmin_mean)) +stat_summary(fun.y = "mean", geom = "point", col = "red", size = 2) +
  theme(axis.text = element_text(size = 14, face = "bold"), axis.title = element_text(size = 14, face = "bold")) +
  geom_smooth(method = "lm")

```

Slope = 0.017, p = 0.0069 so statistically significant

```{r}
# regression line for Max temp, year average

res1 = lm(tmax_mean~year, data=clim_year_avg)
summary(res1)
confint(res1, "year", level = 0.95)
ggplot(clim_year_avg, aes(x = year, y = tmax_mean)) +stat_summary(fun.y = "mean", geom = "point", col = "red", size = 2) +
  theme(axis.text = element_text(size = 14, face = "bold"), axis.title = element_text(size = 14, face = "bold")) +
  geom_smooth(method = "lm")
```

slope = 0.02178 F/year, statistically significant 0.0126

```{r}

# regression line for mean year precipitation

res2 = lm(precip_mean~year, data=clim_year_avg)
summary(res2)
confint(res2, "year", level = 0.95)
ggplot(clim_year_avg, aes(x = year, y = precip_mean)) +stat_summary(fun.y = "mean", geom = "point", col = "navy", size = 2) +
  labs(x = "Year", y = "Average Precipitation (cm)", title = "Average Annual Precipitation Levels (cm)")+
  geom_smooth(method = "lm", color = "black") +
  theme_minimal()+
  theme(plot.title = element_text(hjust=0.5), text = element_text(family = "Times New Roman"))

ggsave(here("figs", "avg.precip.png"), width = 8, height = 5)
```

slope = -5.61 x 10-5, p value = 0.2005, not statistically signficant


### We can look at trends with shorter periods of time to see if there is a more significant trend in shorter time
#### going to look at 2000-2020

```{r}
clim_20 <- clim_1950 %>% 
  filter(year >= 2000,
         year != 2021) %>% 
  group_by(year) %>% 
  summarize(tmax_mean=mean(tmax), tmin_mean = mean(tmin), precip_mean=mean(prcp))



ggplot(clim_20, aes(year, tmax_mean))+geom_point()+ labs(y="Mean Annual Daily Maximum Temperature (degrees F)", x="Date")

ggplot(dat = clim_20, aes(x = year, y = tmin_mean)) + geom_point()+ labs( y= "Mean Annual Daily Minimum Temperature")

ggplot(data =clim_20, aes(x=year, y = precip_mean)) + geom_point() + labs(y= "Mean Annual Daily Precipitation")
```


#Finding regression for 1980-2020 (2000-2020)
##tmin
```{r}
# regression line for min temp, year average

res = lm(tmin_mean~year, data=clim_20)
summary(res)
confint(res, "year", level = 0.95)
ggplot(clim_20, aes(x = year, y = tmin_mean)) +stat_summary(fun.y = "mean", geom = "point", col = "red", size = 2) +
  theme(axis.text = element_text(size = 14, face = "bold"), axis.title = element_text(size = 14, face = "bold")) +
  geom_smooth(method = "lm")
```

##max
```{r}

res1 = lm(tmax_mean~year, data=clim_20)
summary(res1)
confint(res1, "year", level = 0.95)
ggplot(clim_20, aes(x = year, y = tmax_mean)) +stat_summary(fun.y = "mean", geom = "point", col = "red", size = 2) +
  theme(axis.text = element_text(size = 14, face = "bold"), axis.title = element_text(size = 14, face = "bold")) +
  geom_smooth(method = "lm")
```


## Precipitation
```{r}
res2 = lm(precip_mean~year, data=clim_20)
summary(res2)
confint(res2, "year", level = 0.95)
ggplot(clim_20, aes(x = year, y = precip_mean)) +stat_summary(fun.y = "mean", geom = "point", col = "red", size = 2) +
  theme(axis.text = element_text(size = 14, face = "bold"), axis.title = element_text(size = 14, face = "bold")) +
  geom_smooth(method = "lm")
```


Regression line assumes linear relationship, which we do not know, so we need to use Mann-Kendall test
- Mann- Kendall test is used to determine whether a time series has a monotonic upward or downward trend. it does not require that the data be normally distributed or linear
H-null = no correlation, no trend (between time & temp / precipitation)
if S > 0 then later observations in the time series tend to be larger than those that appear earlier in the time series
if S <- then later observations in the time series tend to be smaller than those that appear earlier in time series
```{r}
library(Kendall)
tmin_MK<- MannKendall(clim_year_avg$tmin_mean)
summary(tmin_MK)

tmin_table <- broom::tidy(tmin_MK) 

kable(tmin_table, caption = "Table 1.0 Mann-Kendall Results of Yearly Average Minimum Temperature (1950-2020)",
  col.names = c("T-Statistic", "p-value", "Kendall Score", "Denominator", "Variable Kendall Score"), align = 'c') %>% 
  kable_styling() %>% 
  kable_classic_2(full_width = F, html_font = "Times New Roman")



```

```{r}
tmax_MK <-MannKendall(clim_year_avg$tmax_mean)
tmax_table <- broom::tidy(tmax_MK)

kable(tmax_table, caption = "Table 2.0 Mann-Kendall Results of Yearly Average Maximum Temperature (1950-2020)",
  col.names = c("T-Statistic", "p-value", "Kendall Score", "Denominator", "Variable Kendall Score"), align = 'c') %>% 
  kable_styling() %>% 
  kable_classic_2(full_width = F, html_font = "Times New Roman")

```

```{r}
precip_MK <- MannKendall(clim_year_avg$precip_mean)
precip_table <-broom::tidy(precip_MK)

kable(precip_table, caption = "Table 3.0 Mann-Kendall Results of Yearly Average Precipitation Levels (1950-2020)",
  col.names = c("T-Statistic", "p-value", "Kendall Score", "Denominator", "Variable Kendall Score"), align = 'c') %>% 
  kable_styling() %>% 
  kable_classic_2(full_width = F, html_font = "Times New Roman")
```



```{r}
#time series graph of yearly daily avg of minimum temp (1950-2020)
ggplot(data = clim_year_avg, aes(x=year, y=tmin_mean)) + geom_line()+
  stat_smooth(method = lm)
       
#time series graph of yearly daily avg of max temp (1950-2020)
ggplot(data = clim_year_avg, aes(x=year, y=tmax_mean)) + geom_line()+
  stat_smooth(method = lm)

#time series graph of yearly daily avg of precip  (1950-2020)
ggplot(data = clim_year_avg, aes(x=year, y=precip_mean)) + geom_line()+
  stat_smooth(method = lm)
```

two-sided t-test
```{r}
# splitting the minimum data into two time series - found significant like this
t.test(subset(clim_year_avg$tmin_mean, clim_year_avg$year %in% 1950:1999), subset(clim_year_avg$tmin_mean, clim_year_avg$year %in% 2000:2020))

# splitting the max data into two time series - significant
t.test(subset(clim_year_avg$tmax_mean, clim_year_avg$year %in% 1950:1999), subset(clim_year_avg$tmax_mean, clim_year_avg$year %in% 2000:2020))

# splitting the precip data into two time series - NOT significant
t.test(subset(clim_year_avg$precip_mean, clim_year_avg$year %in% 1950:1999), subset(clim_year_avg$precip_mean, clim_year_avg$year %in% 2000:2020))
```





### More Data Visualizations

#### want to look at it by monthly averages
```{r}
clim_month_avg <- clim_1950 %>% 
  mutate(month = month(date),
         date = date(date)) %>%
  group_by(year, month) %>% 
  summarize(precip_mean = mean(prcp), tmin_mean=mean(tmin), tmax_mean=mean(tmax))

ggplot(data = clim_month_avg, aes(x = year, y = precip_mean)) + geom_line()+
  facet_wrap(~month)

```


#### looking at it seasonally
```{r}
clim_season <- clim_month_avg %>% 
  mutate(season = case_when(month %in% c(12, 1, 2) ~ "Winter",
            month %in% c(3,4,5) ~ "Spring",
            month %in% c(6,7,8) ~ "Summer",
            month %in% c(9, 10, 11) ~ "Fall")) %>% 
  filter(year >= 2000,
         year != 2021)


ggplot(data = clim_season, aes(x=year, y = precip_mean)) + geom_line()+
  facet_wrap(~season)+
  stat_smooth(method = lm)
```

```{r}
ggplot(data = clim_season, aes(x=year, y = tmin_mean)) + geom_line()+
  facet_wrap(~season)+
  stat_smooth(method = lm)
```

```{r}
ggplot(data = clim_season, aes(x=year, y = tmax_mean)) + geom_line()+
  facet_wrap(~season)+
  stat_smooth(method = lm)
```



### Looking at Extremes
from 1950-2020
```{r}
clim_1950$freeze = ifelse(clim_1950$tmin <=32,1,0) #if tmin is 32 degrees of less, put a 1. if above 32 degress put a 0
clim_pk <- clim_1950 %>% 
  group_by(year) %>% 
  summarize(ndayfr = sum(freeze))

ggplot(data= clim_pk, aes(x=year, y =ndayfr)) + geom_point()+labs(y="Number of Freezing Days") +
  stat_smooth(method = lm)

```

```{r}
#two sided t test of number of days from 1950-1985, and 1986-2020
tmin_ttest <- t.test(subset(clim_pk$ndayfr, clim_pk$year %in% 1950:1985), subset(clim_pk$ndayfr, clim_pk$year %in% 1986:2020))
tmin_ttest_table <- broom::tidy(tmin_ttest)

kable(tmin_ttest_table, caption = "Table 3.0 Two Sided T-test of Number of Freezing Days") %>% 
  kable_styling() %>% 
  kable_classic_2(full_width = F, html_font = "Times New Roman")
```

```{r}
#mann-kendall of min
freeze_MK <- MannKendall(clim_pk$ndayfr)
summary(freeze_MK)
```


## looking for the hottest day of the year
```{r}
clim_hot <-clim_1950 %>% group_by(year) %>%  summarize(hotday=max(tmax))

ggplot(data = clim_hot, aes(x=year, y = hotday)) + geom_point(color = "navy")+labs(y="Temperature (degrees F)", x = "Year", title = "Hottest Days of the Year (1950-2020)")  +
  stat_smooth(method = lm, color = "black")+
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5), text = element_text(family = "Times New Roman"))

ggsave(here("figs", "hotdays.png"), width = 5, height =3)
```

```{r}
#hot days ttest
hot_test <-15 t.test(subset(clim_hot$hotday, clim_hot$year %in% 1950:1985), subset(clim_hot$hotday, clim_hot$year %in% 1986:2020))

```


```{r}
clim_1950$rain = ifelse(clim_1950$prcp == 0,0,1) #if precip is 0inches then say 0, put a 1. if more than 0 say 1

clim_rain <- clim_1950 %>% 
  group_by(year) %>% 
  summarize(daysrained = sum(rain))

ggplot(data= clim_rain, aes(x=year, y =daysrained)) + geom_point(color = "navy")+labs(title ="Number of Rainy Days over time", y = "Number of Rainy Days", x = "Year")+
  stat_smooth(method = lm, color = "black")+
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5), text = element_text(family = "Times New Roman"))

ggsave(here("figs", "rainydays.png"), width = 5, height = 3)
```


```{r}
#mann-kendall of rain
rain_MK <- MannKendall(clim_rain$daysrained)
summary(rain_MK)

#rain days ttest
t.test(subset(clim_rain$daysrained, clim_hot$year %in% 1950:1985), subset(clim_rain$daysrained, clim_hot$year %in% 1986:2020))

```



Making tables of the # of times that freezing days and rained
```{r}
library(kableExtra)
library(gt)

clim_table <- clim_1950 %>% group_by(year) %>% 
  summarize(Rain = sum(rain), Freezing = sum(freeze)) %>% 
  mutate_all(~replace(., is.na(.),0)) %>% 
  mutate(Year = year) %>% 
  select(Year, Rain, Freezing)

table_a <- clim_table %>% 
  gt %>%
  tab_header(
    title = "Frequency of Climate Extremes",
    subtitle = "Number of days each year that it rained & the minimum temperature dropped below freezing"
  ) %>% 
  tab_options(table.width = pct(40))



```













