---
title: "snow_pack"
author: "Danielle Sclafani"
date: "4/18/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(chron)
library(here)
library(janitor)
```

#### Reading in the data
```{r}
snow_data <- read_csv(here("data", "bishop_snow.csv")) %>% 
  clean_names()
```



```{r}
#missing a big chunk of data between 1898, going to choose data between 1950-2021, because missing large chunk before 1950

snow_1950 <- snow_data %>% 
  mutate(year = year(date)) %>% 
  filter(year >= 1950,
         year != 2021) #removed 2021 because do not have a full year of data, was skewing data

# need to fill the rows that are N/As
fillrow = which(is.na(snow_1950$snow))
fillrow
snow_1950$snow[fillrow]=0

```



```{r}

ggplot(snow_1950, aes(date, snow_1950$snow))+geom_line()+ labs(y="Snow", x="Date")

```




The trend has a lot of seasonality and is very noisy so going to try annual averages - adding a trend line to the tmin and tmax
```{r, fig.cap="Figure 1.0 Initial Data Visualization of Annual Average Daily Minimum and Maximum temperatures from 1950-2020"}
snow_year_avg <-
  snow_1950 %>% 
  group_by(year) %>% 
  summarize(snow_mean=mean(snow))


a <- ggplot(data = snow_year_avg, aes(x = year, y = snow_mean)) + geom_point(color = "skyblue") +
geom_point(data=snow_year_avg, aes(x=year, y = snow_mean), col="cyan4")

#adding a trendline
a = a+stat_smooth(method = "lm", col = "skyblue")
a
a+stat_smooth(data = snow_year_avg, aes(x = year, y = snow_mean), col = "cyan4", method = lm)+theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5), text = element_text(family = "Times New Roman"))

```


## Finding the slope of the trendlines
```{r}

# regression line for min temp, year average

res = lm(snow_mean~year, data=snow_year_avg)
summary(res)
confint(res, "year", level = 0.95)
ggplot(clim_year_avg, aes(x = year, y = tmin_mean)) +stat_summary(fun.y = "mean", geom = "point", col = "red", size = 2) +
  theme(axis.text = element_text(size = 14, face = "bold"), axis.title = element_text(size = 14, face = "bold")) +
  geom_smooth(method = "lm")

```

Slope = -0.000, not significant


### We can look at trends with shorter periods of time to see if there is a more significant trend in shorter time
#### going to look at 2000-2020

```{r}
snow_20 <- snow_1950 %>% 
  filter(year >= 2000,
         year != 2021) %>% 
  group_by(year) %>% 
  summarize(snow_mean=mean(snow))



ggplot(snow_20, aes(year, snow_mean))+geom_point()

```


#Finding regression for 1980-2020 (2000-2020)
##tmin
```{r}
# regression line for min temp, year average

res = lm(snow_mean~year, data=snow_20)
summary(res)
confint(res, "year", level = 0.95)
ggplot(snow_20, aes(x = year, y = snow_mean)) +stat_summary(fun.y = "mean", geom = "point", col = "red", size = 2) +
  theme(axis.text = element_text(size = 14, face = "bold"), axis.title = element_text(size = 14, face = "bold")) +
  geom_smooth(method = "lm")
```


Regression line assumes linear relationship, which we do not know, so we need to use Mann-Kendall test
- Mann- Kendall test is used to determine whether a time series has a monotonic upward or downward trend. it does not require that the data be normally distributed or linear
H-null = no correlation, no trend (between time & temp / precipitation)
if S > 0 then later observations in the time series tend to be larger than those that appear earlier in the time series
if S <- then later observations in the time series tend to be smaller than those that appear earlier in time series
```{r}
library(Kendall)
snow_MK<- MannKendall(snow_year_avg$snow_mean)
summary(snow_MK)

tmin_table <- broom::tidy(tmin_MK) 

```


```{r}
#time series graph of yearly daily avg of minimum temp (1950-2020)
ggplot(data = snow_year_avg, aes(x=year, y=snow_mean)) + geom_line()+
  stat_smooth(method = lm)

```

two-sided t-test
```{r}
# splitting the minimum data into two time series - found significant like this
t.test(subset(snow_year_avg$snow_mean, snow_year_avg$year %in% 1950:1999), subset(snow_year_avg$snow_mean, snow_year_avg$year %in% 2000:2020))

```




